---
title: "p8105_final_project"
author: "Xinwei Han (xh2601), Xinhui Lin (xl3054), Ruohan Lyu (rl3610), Xinyin Miao (xm2356), Fenglin Xie (fx2212)"
date: "2025-12-06"
output: github_document
---

load packages
```{r}
library(tidyverse)
library(hms)
library(lubridate)
library(viridis)
library(ggthemes)
library(scales)
```

Data cleaning: remove incorrect records based on call & response time, split date and time
```{r}
park_ranger = read_csv("data/Urban_Park_Ranger_Animal_Condition_Response_20251103.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    date_and_time_of_initial_call = mdy_hm(date_and_time_of_initial_call),
    date_and_time_of_ranger_response = mdy_hm(date_and_time_of_ranger_response),
    
    call_date = as.Date(date_and_time_of_initial_call),
    call_time = as_hms(date_and_time_of_initial_call),
    response_date = as.Date(date_and_time_of_ranger_response),
    response_time = as_hms(date_and_time_of_ranger_response)  ## split date and time
  ) %>% 
  filter(date_and_time_of_ranger_response >= date_and_time_of_initial_call)
```

Reclassify animal classes into 7 categories
```{r}
park_ranger_new = park_ranger %>% 
  mutate(
    animal_class_new = case_when(
      species_description %in% c(
        "Atlantic Canary","Bird (Unknown)","Budgerigar","Budgerigar Parakeet",
        "Chicken","Chukar","Cockatiel","Domestic Dove","Domestic Duck", "Domestic duck",
        "Domestic Goose","Domestic Goose Hybrid","Domestic Quail","Domestic Turkey",
        "Domestic Waterfowl","Fancy Dove","Guinea Fowl","Guineafowl","Japanese Quail",
        "Khaki Campbell","Muscovy Duck","Rock Dove",
        "Parakeet (Unknown)"
      ) ~ "Birds / Raptors",
      
      species_description %in% c(
        "Domestic Rabbit","Domestic Ferret","Fancy Rat","Gerbil",
        "Guinea Pig", "Guniea Pig", "Guinea pig", "Hamster"
      ) ~ "Small Mammals",
      
      species_description %in% c(
        "Dog","Pitt Bull Mix","Cat","Goat","Cattle","Pig"
      ) ~ "Large Mammals",
      
      species_description %in% c(
        "Animal (Unknown)","Honey Bee", "Unknown","N/A"
      ) ~ "Others",
      
      animal_class %in% c(
        "[\"Birds\"]", "[\"Raptors\"]", "Birds", "Raptors", 
        "Raptors, Small Mammals-RVS"
      ) ~ "Birds / Raptors",
            
      animal_class %in% c(
        "[\"Small Mammals-non RVS\"]", "[\"Small Mammals-RVS\"]",
        "Small Mammals-non RVS", "Small Mammals-RVS"
      ) ~ "Small Mammals",
      
      animal_class %in% c(
        "Coyotes", "Deer", "[\"Coyotes\"]", "[\"Deer\"]"
      ) ~ "Large Mammals",

      animal_class %in% c(
        "[\"Marine Mammals-whales, Dolphin\"]", "[\"Marine Mammals-seals only\"]",
        "Marine Mammals-seals only", "Marine Mammals-whales,  Dolphin", 
        "Marine Mammals-whales, Dolphin"
      ) ~ "Marine Mammals",
            
      animal_class %in% c(
        "[\"Fish-numerous quantity\"]", "[\"Non Native Fish-(invasive)\"]", 
        "Fish-numerous quantity", "Fish-numerous quantity;#Terrestrial Reptile or Amphibian", 
        "Non Native Fish-(invasive)"
      ) ~ "Fish",
      
      animal_class %in% c(
        "[\"Marine Reptiles\"]", "[\"Terrestrial Reptile or Amphibian\"]", 
        "Marine Reptiles", "Terrestrial Reptile or Amphibian", 
        "Terrestrial Reptile or Amphibian, Fish-numerous quantity"
      ) ~ "Reptiles / Amphibians",

      animal_class %in% c(
        "[\"Rare, Endangered, Dangerous\"]", "Rare,  Endangered,  Dangerous",
        "Rare, Endangered, Dangerous"
      ) ~ "Others"
    )
  )
```

```{r}
table(park_ranger_new$animal_class_new)
```


# TASK 1: Monthly trends by animal class using the new classification

```{r}
monthly_animal_trends = park_ranger_new |> 
  mutate(
    call_month = month(date_and_time_of_initial_call, label = TRUE, abbr = TRUE)
  ) |> 
  group_by(call_month, animal_class_new) |> 
  summarise(case_count = n(), .groups = "drop") |> 
  # Ensure all months are represented for each animal class
  complete(call_month, animal_class_new, fill = list(case_count = 0))
```


# Plot 1: Monthly trends by animal class (new classification)

```{r}
mon_case_plot = monthly_animal_trends |> 
  ggplot(aes(x = call_month, y = case_count, group = animal_class_new, color = animal_class_new)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Monthly Animal Response Cases by Animal Class (New Classification)",
    subtitle = "Urban Park Ranger Animal Condition Response",
    x = "Month",
    y = "Number of Cases",
    color = "Animal Class"
  ) +
  theme_fivethirtyeight() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    legend.text = element_text(size = 8)
  ) +
  scale_color_viridis_d()

print(mon_case_plot)
```

Chart Interpretation:

This line chart displays the monthly distribution of animal response cases across 7 reclassified animal categories throughout the year.

Key Findings:

1. Birds and raptors demonstrate a notable increase in cases beginning in February, reaching a peak in June, and declining to their lowest levels by November, possibly reflecting migratory patterns, nesting seasons, and reduced activity during colder months. 

2. Small mammals exhibit three distinct peaks in case numbers during April, August, and October, which may suggest potential associations with breeding cycles, seasonal resource availability, or periodic human-wildlife interactions in urban environments.

3. Large mammals maintain relatively consistent case numbers between 70 and 80 throughout the year, indicating potentially stable population dynamics or continuous human encounters despite seasonal variations.

4. Reptiles and amphibians show generally low and stable case numbers with a moderate peak around 100 in June, which might be linked to temperature-dependent activity patterns and increased visibility during warmer conditions.

5. Marine mammals and other categories remain consistently minimal with case counts in the single digits, possibly due to their limited presence in urban park ecosystems or lower detection rates.


# TASK 2: Heatmap of cases by month and hour (using cleaned data)

```{r}
hourly_monthly_heatmap = park_ranger_new |> 
  mutate(
    call_month = month(date_and_time_of_initial_call, label = TRUE, abbr = TRUE),
    call_hour = hour(date_and_time_of_initial_call)
  ) |> 
  filter(!is.na(call_month) & !is.na(call_hour)) |> 
  group_by(call_month, call_hour) |> 
  summarise(case_count = n(), .groups = "drop") |> 
  complete(call_month, call_hour, fill = list(case_count = 0))
```

Create custom color gradient with more detailed breaks

```{r}
max_cases = max(hourly_monthly_heatmap$case_count)
color_breaks = c(0, 1, 5, 10, 20, 50, max_cases)
```

Define red color palette

```{r}
red_palette = colorRampPalette(c("#FFF5F0", "#FEE0D2", "#FCBBA1", "#FC9272", "#FB6A4A", "#EF3B2C", "#CB181D", "#99000D"))(length(color_breaks))
```

# Plot 2: Heatmap of cases by month and hour

```{r}
case_heatmap = hourly_monthly_heatmap |>
  ggplot(aes(x = call_hour, y = call_month, fill = case_count)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradientn(
    name = "Case Count",
    colors = red_palette,
    values = rescale(color_breaks),
    breaks = color_breaks,
    guide = guide_colorbar(
      barwidth = 30,
      barheight = 0.8,
      direction = "horizontal",
      title.position = "top"
    )
  ) +
  scale_x_continuous(
    breaks = seq(0, 23, by = 2),
    labels = c("12AM", "2AM", "4AM", "6AM", "8AM", "10AM", 
               "12PM", "2PM", "4PM", "6PM", "8PM", "10PM")
  ) +
  labs(
    title = "Animal Response Cases: Detailed Hourly and Monthly Distribution",
    subtitle = "Enhanced color gradient reveals subtle patterns in case distribution",
    x = "Hour of Day",
    y = "Month"
  ) +
  theme_fivethirtyeight() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 9),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    legend.position = "bottom",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

plot(case_heatmap)

```

## Time-Month Heatmap Analysis:

Case distribution primarily concentrates between 8:00 and 18:00, aligning with peak park visitation and diurnal animal activity. Summer exhibits extended evening activity likely due to increased daylight and recreational use, while winter shows contracted midday patterns corresponding to reduced daylight availability. Minimal overnight reporting may reflect limited detection capabilities during low-human-activity periods.

## Human Intervention Recommendations:

Enhanced ranger coverage during summer afternoons could address elevated case frequencies across multiple species. Early spring public education initiatives might mitigate conflicts by raising awareness of seasonal animal behaviors. Rescue organizations should anticipate increased warm-weather intake while utilizing winter for strategic planning. Visitor vigilance during peak seasons and prompt daylight reporting could improve response effectiveness, supplemented by community monitoring programs and coordinated rehabilitation networks.


# TASK 7

```{r}
park_model = park_ranger_new |>
  rename(num_animals = matches("animals")) |>
  mutate(
    duration_hours      = as.numeric(duration_of_response),
    num_animals         = as.numeric(num_animals),
    final_ranger_action = as.factor(final_ranger_action),
    animal_condition    = as.factor(animal_condition)
  ) |>
  filter(
    !is.na(duration_hours),
    duration_hours > 0,
    !is.na(final_ranger_action),
    !is.na(animal_condition),
    !is.na(num_animals),
    num_animals >= 0
  )

```

# Violin plot: duration vs final_ranger_action
```{r}
ggplot(park_model, aes(x = final_ranger_action, y = duration_hours)) +
  geom_violin(trim = FALSE, fill = "skyblue", color = "grey30") +
  coord_flip() +
  labs(
    x = "Final ranger action",
    y = "Duration of response (hours)",
    title = "Violin plot of response duration by final ranger action"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )
```

# Boxplot: duration vs final_ranger_action
```{r}
ggplot(park_model, aes(x = final_ranger_action, y = duration_hours)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "lightgreen") +
  coord_flip() +
  labs(
    x = "Final ranger action",
    y = "Duration of response (hours)",
    title = "Boxplot of response duration by final ranger action"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )
```

# Violin plot: duration vs animal_condition
```{r}
ggplot(park_model, aes(x = animal_condition, y = duration_hours)) +
  geom_violin(trim = FALSE, fill = "skyblue", color = "grey30") +
  coord_flip() +
  labs(
    x = "Animal condition",
    y = "Duration of response (hours)",
    title = "Violin plot of response duration by animal condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )

```

# Boxplot: duration vs animal_condition
```{r}
ggplot(park_model, aes(x = animal_condition, y = duration_hours)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "lightgreen") +
  coord_flip() +
  labs(
    x = "Animal condition",
    y = "Duration of response (hours)",
    title = "Boxplot of response duration by animal condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )

```

# Density plot

```{r}
park_model2 = park_model |>
  mutate(
    action_group = if_else(
      final_ranger_action == "Unfounded",
      "Unfounded",
      "Other actions"
    )
  )
```

```{r}
duration_mean = park_model2 |>
  group_by(action_group) |>
  summarise(
    mean_duration = mean(duration_hours, na.rm = TRUE),
    .groups = "drop"
  )
duration_mean
```

```{r}
ggplot(park_model2, aes(x = duration_hours,
                        fill  = action_group,
                        color = action_group)) +
  geom_density(alpha = 0.3, adjust = 1, linewidth = 1) +
  
  geom_vline(data = duration_mean,
             aes(xintercept = mean_duration, color = action_group),
             linewidth = 1.2) +
  
  labs(
    title = "Density of response duration by action group",
    x = "Duration of response (hours)",
    y = "Density",
    fill  = "Action group",
    color = "Action group"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```


```{r}
ggplot(park_model2,
       aes(x = duration_hours, fill = action_group, color = action_group)) +
  geom_density(alpha = 0.4, adjust = 1, linewidth = 0.7) +
  coord_cartesian(xlim = c(0, 4)) +   # 只看 0–4 小时
  labs(
    title = "Density of response duration (zoomed to 0–4 hours)",
    x = "Duration of response (hours)",
    y = "Density",
    fill  = "Action group",
    color = "Action group"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```


```{r}
ggplot(park_model2,
       aes(x = duration_hours, fill = action_group)) +
  geom_density(alpha = 0.6, adjust = 1, linewidth = 0.6, color = NA) +
  coord_cartesian(xlim = c(0, 4)) +
  facet_wrap(~ action_group, ncol = 1, scales = "free_y") +
  labs(
    title = "Response duration density by action group (0–4 hours)",
    x = "Duration of response (hours)",
    y = "Density"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```

## log 变换
```{r}
park_model2_log = park_model2 |>
  filter(duration_hours > 0) |>
  mutate(duration_log = log1p(duration_hours))  # log(1 + x)

ggplot(park_model2_log,
       aes(x = duration_log, fill = action_group, color = action_group)) +
  geom_density(alpha = 0.4, adjust = 1, linewidth = 0.7) +
  labs(
    title = "Density of log-transformed response duration",
    x = "log(1 + duration of response, hours)",
    y = "Density",
    fill  = "Action group",
    color = "Action group"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```

# ANOVA: Duration ~ Final Action + Condition + # of Animals

```{r}

anova_fit = aov(
  duration_hours ~ final_ranger_action * animal_condition + num_animals,
  data = park_model
)

summary(anova_fit)

```

## log
```{r}
park_model2 = park_model |>
  filter(duration_hours > 0) |>
  mutate(log_duration = log(duration_hours))

anova_fit_log = aov(
  log_duration ~ final_ranger_action * animal_condition + num_animals,
  data = park_model2
)

summary(anova_fit_log)

par(mfrow = c(1, 2))
plot(anova_fit_log, which = 1)  # Residuals vs Fitted
plot(anova_fit_log, which = 2)  # Normal Q-Q
par(mfrow = c(1, 1))

```


# Task 8

```{r}
park_model4 <- park_model |>
  filter(!is.na(animal_condition)) |>
  filter(animal_condition %in% c("Healthy", "Unhealthy", "Injured", "DOA"))

```

```{r}
plot_condition_bar <- function(data, cond) {
  data |>
    filter(animal_condition == cond) |>
    ggplot(aes(x = final_ranger_action)) +
    geom_bar(fill = "#4F81BD", alpha = 0.85) +
    coord_flip() +
    labs(
      title = paste("Final Ranger Action for", cond),
      x = "Final Ranger Action",
      y = "Number of Cases"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.y = element_text(size = 8)
    )
}
```


```{r}
library(patchwork)

p1 <- plot_condition_bar(park_model4, "Healthy")
p2 <- plot_condition_bar(park_model4, "Unhealthy")
p3 <- plot_condition_bar(park_model4, "Injured")
p4 <- plot_condition_bar(park_model4, "DOA")

(p1 | p2) /
(p3 | p4)

```
















