---
title: "Project Report"
author: Xinwei Han (xh2601), Xinhui Lin (xl3054), Ruohan Lyu (rl3610), Xinyin Miao (xm2356), Fenglin Xie (fx2212)
date: "2025-12-06"
output: html_document
---

```{r, message = FALSE}
library(tidyverse)
library(hms)
library(lubridate)
library(viridis)
library(ggthemes)
library(scales)
library(glmnet)
library(patchwork)
```

# Tiny Paws, Big Rescue: 
## A Data Story of Seasonal and Spatial Animal Saving

## Motivation: Provide an overview of the project goals and motivation.
During the 2025 NYC Tree Count, one of our team members observed a raccoon hiding inside a tree cavity in Morningside Park. Park rangers noted that animals living in urban parks are often stressed or displaced by human activity, yet these encounters also give citizens opportunities to notice injuries, abnormal behavior, or animals in distress. These experiences highlight the importance of citizen science in helping residents recognize that cities serve as shared habitats for both humans and wildlife, and in encouraging community involvement in reporting and supporting animals that may need assistance.

Animals play an essential role in maintaining ecological balance, yet those living in dense urban environments like New York City face unique challenges, including traffic, pollution, habitat fragmentation, and frequent interactions with humans. By analyzing the Urban Park Ranger Animal Response dataset, we aim to identify when and where animals most frequently require help, ultimately raising public awareness and supporting timely rescue and conservation efforts.

To achieve this, we are developing an interactive, data-driven website that allows visitors to explore animal rescue patterns through a variety of visualizations. For instance, the site includes interactive maps that display the geographic locations of animal aid incidents, enabling users to identify high-risk areas and examine the number of reported cases in parks of interest. By showing where different animal species are typically found, when incidents occur most frequently, and how park rangers respond, our platform aims to connect the public with the city’s urban wildlife and promote informed, community-supported conservation.


# Related work: Anything that inspired you, such as a paper, a web site, or something we discussed in class.
Our project is inspired by a range of citizen science and urban ecology initiatives that highlight how collaboration between the public and park agencies can improve wildlife protection in cities. For example, large-scale volunteer programs such as the NYC Tree Census (led by NYC Parks and NYC TreesCount!) demonstrate how citizen science can successfully mobilize thousands of residents to monitor environmental conditions across the city. Similar community-based wildlife monitoring efforts, such as iNaturalist, eBird, and the Urban Wildlife Information Network (UWIN), show that public participation can generate high-quality ecological data and increase awareness of human–wildlife interactions.

# Initial questions: What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?

We originally planned to analyze the associations between animal conditions, locations, and time, both across the year and throughout the day. Additional analyses will examine factors that may influence response time, such as park location, animal condition or number, and temporal variables like time of day or frequency within a week or month. Examples of visualizations include a heatmap illustrating the density of injured animals by month and time of day, as well as violin plots grouped by boroughs to show the distribution of response times across regions. We will also conduct statistical tests, such as ANOVA, to evaluate relationships among variables, including borough, response time, animal condition, and time factors.

# Data: Source, scraping method, cleaning, etc.

## Data Source

Our primary dataset is provided by the Department of Parks and Recreation (DPR) and is made publicly available through the NYC OpenData portal [(Urban Park Ranger Animal Condition Response)](https://data.cityofnewyork.us/Environment/Urban-Park-Ranger-Animal-Condition-Response/fuhs-xmg2/about_data). This dataset contains 6,385 records of animal rescue requests, including detailed information such as the date and time of each call, ranger response time, park name, specific location descriptions, animal species, and the condition of the animals involved.

One essential component of the website visualization is displaying the number of total cases reported to park rangers for each park. The Park Ranger dataset does not contain geographic coordinates, but it provides the park name or the location associated with each case. To map these cases spatially, we used the [Parks Properties](https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve/about_data) dataset released by the Department of Parks and Recreation (DPR) on NYC OpenData. This dataset contains 2,054 parks and corresponding geographic coordinates in New York City across five boroughs. Coordinates of the 5 boroughs are from the [Borough Boundaries](https://data.cityofnewyork.us/City-Government/Borough-Boundaries/gthc-hcne/about_data) dataset released by the Department of City Planning (DCP) on NYC OpenData.

## Data cleaning process

- Split date and time

- Remove incorrect records based on call & response time

- Rename variable `duration_of_response` to `duration_of_action`

```{r, echo = FALSE, message = FALSE}
park_ranger = read_csv("data/Urban_Park_Ranger_Animal_Condition_Response_20251103.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    date_and_time_of_initial_call = mdy_hm(date_and_time_of_initial_call),
    date_and_time_of_ranger_response = mdy_hm(date_and_time_of_ranger_response),
    
    call_date = as.Date(date_and_time_of_initial_call),
    call_time = as_hms(date_and_time_of_initial_call),
    response_date = as.Date(date_and_time_of_ranger_response),
    response_time = as_hms(date_and_time_of_ranger_response)
  ) %>% 
  filter(date_and_time_of_ranger_response >= date_and_time_of_initial_call) %>% 
  rename(duration_of_action = duration_of_response)
```

- Reclassify animal classes into 7 categories: Birds / Raptors, Small Mammals, Large Mammals, Marine Mammals, Reptiles / Amphibians, Fish, Others

- Factor categorical variables & set reference level

```{r}
park_ranger_new = park_ranger %>% 
  mutate(
    animal_class_new = case_when(
      animal_class %in% c(
        "[\"Birds\"]", "[\"Raptors\"]", "Birds", "Raptors", 
        "Raptors, Small Mammals-RVS"
      ) ~ "Birds / Raptors",
            
      animal_class %in% c(
        "[\"Small Mammals-non RVS\"]", "[\"Small Mammals-RVS\"]",
        "Small Mammals-non RVS", "Small Mammals-RVS"
      ) ~ "Small Mammals",
      
      animal_class %in% c(
        "Coyotes", "Deer", "[\"Coyotes\"]", "[\"Deer\"]"
      ) ~ "Large Mammals",

      animal_class %in% c(
        "[\"Marine Mammals-whales, Dolphin\"]", "[\"Marine Mammals-seals only\"]",
        "Marine Mammals-seals only", "Marine Mammals-whales,  Dolphin", 
        "Marine Mammals-whales, Dolphin"
      ) ~ "Marine Mammals",
            
      animal_class %in% c(
        "[\"Fish-numerous quantity\"]", "[\"Non Native Fish-(invasive)\"]", 
        "Fish-numerous quantity", "Fish-numerous quantity;#Terrestrial Reptile or Amphibian", 
        "Non Native Fish-(invasive)"
      ) ~ "Fish",
      
      animal_class %in% c(
        "[\"Marine Reptiles\"]", "[\"Terrestrial Reptile or Amphibian\"]", 
        "Marine Reptiles", "Terrestrial Reptile or Amphibian", 
        "Terrestrial Reptile or Amphibian, Fish-numerous quantity"
      ) ~ "Reptiles / Amphibians",

      animal_class %in% c(
        "[\"Rare, Endangered, Dangerous\"]", "Rare,  Endangered,  Dangerous",
        "Rare, Endangered, Dangerous"
      ) ~ "Others",
      
      species_description %in% c(
        "Atlantic Canary","Bird (Unknown)","Budgerigar","Budgerigar Parakeet",
        "Chicken","Chukar","Cockatiel","Domestic Dove","Domestic Duck", "Domestic duck",
        "Domestic Goose","Domestic Goose Hybrid","Domestic Quail","Domestic Turkey",
        "Domestic Waterfowl","Fancy Dove","Guinea Fowl","Guineafowl","Japanese Quail",
        "Khaki Campbell","Muscovy Duck","Rock Dove",
        "Parakeet (Unknown)"
      ) ~ "Birds / Raptors",
      
      species_description %in% c(
        "Domestic Rabbit","Domestic Ferret","Fancy Rat","Gerbil",
        "Guinea Pig", "Guniea Pig", "Guinea pig", "Hamster"
      ) ~ "Small Mammals",
      
      species_description %in% c(
        "Dog","Pitt Bull Mix","Cat","Goat","Cattle","Pig"
      ) ~ "Large Mammals",
      
      species_description %in% c(
        "Animal (Unknown)","Honey Bee", "Unknown","N/A"
      ) ~ "Others"),
    
    animal_class_new = factor(animal_class_new, 
                              levels = c("Birds / Raptors", "Small Mammals", "Large Mammals", 
                                         "Marine Mammals", "Reptiles / Amphibians", "Fish", "Others")),
    
    borough = factor(borough, levels = c("Manhattan", "Brooklyn", "Queens", "Bronx", "Staten Island")),
    
    call_source = factor(call_source, levels = c("Public", "Central", "Employee", 
                                                 "Conservancies/\"Friends of\" Groups", 
                                                 "Observed by Ranger", "WBF", "WINORR", "Other")),
    
    species_status = ifelse(species_status == "N/A" | is.na(species_status), 
                            "Unknown", species_status),
    species_status = factor(species_status, levels = c("Native", "Invasive", "Domestic", 
                                                       "Exotic", "Unknown")),
    
    animal_condition = ifelse(animal_condition == "N/A" | is.na(animal_condition), 
                              "Unknown", animal_condition),
    animal_condition = factor(animal_condition, levels = c("Healthy", "Unhealthy", "Injured", 
                                                           "DOA", "Unknown")),
    
    adult   = if_else(str_detect(age, "Adult"), 1L, 0L, missing = 0L),
    juvenile = if_else(str_detect(age, "Juvenile"), 1L, 0L, missing = 0L),
    infant  = if_else(str_detect(age, "Infant"), 1L, 0L, missing = 0L)
  )
```

## Matching Process for Park Properties

fter comparing the **Park Ranger** and **Parks Properties** datasets, we found that many park names in Park Ranger (*property*) did not match the official park names (*signname*) in Parks Properties Specifically, 208 out of 703 unique *property* values had no exact match in *signname*. To recover as many matches as possible, we applied the Jaro–Winkler algorithm from the fuzzyjoin package to perform inexact matching on all unique park names.

We selected Jaro–Winkler algorithm because it is designed for short strings such as names, and it places higher weight on common prefixes, which is useful for matching park names that share the same beginning but differ due to abbreviations, typos, or partial truncation (e.g., “Prospect Pk” vs. “Prospect Park”). 

All fuzzy-matched pairs with similarity scores less than 1 were then manually reviewed by three members of the group, and each match was checked independently by two reviewers to ensure accuracy and inter-rater reliability. After this process, 35 unique property values remained unmatched, typically because they referred to street names (e.g., “Green Street”), or parks that exist but are not included in the Parks Properties dataset (e.g., “Mannahatta Park”). The final map includes 6,118 reported cases across 546 unique parks. Using Jaro–Winkler with a maximum distance threshold of 0.25 resulted in an overall matching accuracy of 43.75% among the 208 originally unmatched values.

## Cleaned dataset

The cleaned data includes the following variables:

`date_and_time_of_initial_call`: The date and time of the initial request for animal rescue

`date_and_time_of_ranger_response`: The date and time when the Ranger responded to the request (/arrived at the site)xxxxxxxxxx

`borough`: Borough in which request for assistance was (Manhattan, Brooklyn, Queens, Bronx, Staten Island)

`property`: Name of location of rescue

`location`: Specific cross street of location of rescue

`species_description`: The species being rescued

`call_source`: Location of initial call

`species_status`: The species status (Native, Invasive, Domestic, Exotic, Unknown)

`animal_condition`: The status of the animal reported by initial request

`duration_of_action`: A numerical value representing the amount of time spent responding upon arrival xxxxxxx

`age`: The original age column that contains age of each animal xxxxxxxx

`animal_class`: The original column describing type of species of animal

`x311sr_number`: Service request number generated by the 311 system

`final_ranger_action`: Final action outcome of responding Ranger

`number_of_animals`: Number of animals attended toxxxxxxxx

`pep_response`: An indicator variable representing if Parks Enforcement Patrol was called

`animal_monitored`: An indicator variable representing if the animal was monitored during visit

`rehabilitator`: Name of a rehabilitator if animal was taken to one

`hours_spent_monitoring`: Indicates how many hours were spent monitoring the animal

`police_response`: An indicator variable representing if Police were called

`esu_response`: An indicator variable representing if ESU were called

`acc_intake_number`: Reference number if the animal was sent to ACC

`call_date`: Date of the initial request for animal rescue

`call_time`: Time of the initial request for animal rescue

`response_date`: Date when the Ranger responded to the request (/arrived at the site)xxxxxxxxxx

`response_time`: Time when the Ranger responded to the request (/arrived at the site)xxxxxxxxxx

`animal_class_new`: Reclassification of `animal_class` into 7 categories (Birds / Raptors, Small Mammals, Large Mammals, Marine Mammals, Reptiles / Amphibians, Fish, Others)

`adult`, `juvenile`, `infant`: An indicator variable representing the age of the animal being rescued


# Exploratory analysis: Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

## Response Duration

We would like to explore how response duration, which we define as the duration between the initial request and the time that Rangers arrived at the site (`date_and_time_of_ranger_response - date_and_time_of_initial_call`), is associated with different incident characteristics.

```{r}
ranger_response_hr = park_ranger_new %>% 
  mutate(day_diff = as.numeric(difftime(response_date, call_date, units = "days")),
         time_diff_hrs = as.numeric(difftime(response_time, call_time, units = "hours")),
         response_hrs = time_diff_hrs + 24 * day_diff,
         call_hour = hour(call_time))

ranger_response_hr %>% 
  group_by(call_hour) %>%
  summarise(mean_duration = mean(response_hrs, na.rm = TRUE)) %>% 
  ggplot(aes(x = call_hour, y = mean_duration)) +
  geom_line(linewidth = 0.8) + 
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Average Response Duration by Time of Initial Request",
    x = "Hour of Day (24-hour)",
    y = "Average Response Hours") +
  theme_minimal()
```

This line plot visualizes variation in response duration across the 24-hour day. 

- Cases reported during regular daytime hours (approximately 8:00 - 16:00) have the shortest response duration, typically around 2-4 hours from the time of initial request to the arrival of a Ranger. This pattern suggests that Ranger availability and operational capacity are highest during standard working hours.

- Response duration keeps increasing in the late afternoon and early evening (17:00-22:00), reaching an extended period of long delays in responses, with average response duration approaching 15 hours. This rise likely reflects reduced staffing after 5 PM, and the possibility that non-urgent incidents are deferred until the working hours the next day. 

- Cases reported late at night and in the early morning (approximately 20:00-5:00) also show longer response durations, often exceeding 10 hours, consistent with limited overnight staffing or slower processing of after-hours calls.

- One notable irregularity is the sharp drop in average response duration around 23:00, followed by a spike at 0:00. This pattern may indicate hour misclassification, where some late-night incidents were miscoded as 0:00, combined with small sample sizes or boundary effects at the end of the day. These potential misclassifications should be considered and interpreted with cautions.


```{r, fig.width=14, fig.height=12}
resp_dur_dist = ranger_response_hr %>% 
  mutate(resp_bin = case_when(
    response_hrs < 1 ~ "0-1 hours", 
    response_hrs >= 1 & response_hrs < 4 ~ "1-4 hours",
    response_hrs >= 4 & response_hrs < 24 ~ "4-24 hours",
    response_hrs >= 24 ~ ">24 hours"),
    resp_bin = factor(resp_bin, levels = c("0-1 hours", "1-4 hours", "4-24 hours", ">24 hours"))
  )

resp_dur_dist_plot = function(df, title_suffix) {
  ggplot(df, aes(x = reorder(animal_condition, response_hrs, FUN = median),
                 y = response_hrs)) +
    geom_violin(trim = FALSE, fill = "skyblue", alpha = 0.5) +
    geom_boxplot(width = 0.3, alpha = 0.8) +
    labs(title = title_suffix,
         x = "Animal Condition", y = "Response Duration (hours)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

p1 = resp_dur_dist %>% 
  filter(resp_bin == "0-1 hours") %>% 
  resp_dur_dist_plot("0–1 hours")

p2 = resp_dur_dist %>% 
  filter(resp_bin == "1-4 hours") %>% 
  resp_dur_dist_plot("1–4 hours")

p3 = resp_dur_dist %>% 
  filter(resp_bin == "4-24 hours") %>% 
  resp_dur_dist_plot("4–24 hours")

p4 = resp_dur_dist %>% 
  filter(resp_bin == ">24 hours") %>% 
  resp_dur_dist_plot(">24 hours")

(p1 | p2) / (p3 | p4) + plot_annotation(title ="Distribution of Response Duration")
```

Because response durations are highly right-skewed, the data were split more at the lower end (0–1 and 1–4 hours) where most observations occur, while broader segments were used for longer delays (4–24 and >24 hours). 

The plots show that response duration does not meaningfully differ across animal condition categories. All five groups (Healthy, Injured, Unhealthy, DOA, and Unknown) display highly overlapping distributions with similar central tendencies. The within-condition variability is much larger than any between-condition differences. The long right tails are present across all categories, especially in the >24 hour panel, suggesting that extremely slow responses occur regardless of the reported condition. 

Overall, the figure indicates that the reported animal condition plays only a minimal role. 

```{r}
ranger_model = ranger_response_hr %>%
  mutate(
    call_hour_cat = case_when(call_hour >= 0 & call_hour < 8 ~ "0-8",
                              call_hour >= 8 & call_hour < 16 ~ "8-16",
                              call_hour >= 16 & call_hour < 24 ~ "16-24"),
    call_hour_cat = factor(call_hour_cat, levels = c("8-16", "0-8", "16-24"))
  ) %>% 
  select(response_hrs, borough, call_source, species_status, animal_condition, 
         adult, juvenile, infant, animal_class_new, number_of_animals, call_hour_cat) %>% 
  na.omit()

ranger_model %>% 
  ggplot(aes(x = log(response_hrs), color = borough)) +
  geom_density() +
  labs(title = "Density of Ranger Response Duration by Borough",
       x = "Response Duration (hours)",
       y = "Density",
       color = "Borough") +
  theme_minimal()
```

Because the response-duration distribution is highly right-skewed, with a long tail extending to several hundred hours, we applied a natural log transformation to stabilize the variance and make differences across boroughs more interpretable.

Across boroughs, the density curves show only small differences in response duration. All boroughs have an average of `log(response duration)` less than 0, which means the average of response duration is less than 1 hour, showing fast response duration on average. 

- Manhattan and Brooklyn have slightly right-shifted peaks on the log scale, indicating somewhat slower responses on average compared to others. This might be because these boroughs tend to have higher call volume and more complex environment, which can slow down dispatch and travel time. Manhattan’s dense urban setting and Brooklyn’s large geographic area may also contribute to longer navigation times and resource constraints relative to the other boroughs.

- Queens, the Bronx, and Staten Island have slightly left-shifted peaks on the log scale, indicating somewhat faster responses on average. Operationally, these boroughs may experience less congestion and fewer competing requests, contributing to slightly shorter response duration.

Despite these minor shifts in central tendency, the overall shapes of the distributions are highly similar. Within-borough variation is much greater than between-borough variation.

```{r}
park_model = park_ranger_new |>
  rename(num_animals = matches("animals")) |>
  mutate(
    duration_hours      = as.numeric(duration_of_action),
    num_animals         = as.numeric(num_animals)
  ) |>
  filter(
    !is.na(duration_hours),
    duration_hours > 0,
    !is.na(final_ranger_action),
    !is.na(animal_condition),
    !is.na(num_animals),
    num_animals >= 0
  )

```

## Response Duration by Final Ranger Action
```{r, fig.width=7, fig.height=5, fig.align = "center"}
action_order <- park_model |>
  group_by(final_ranger_action) |>
  summarise(med = median(duration_hours, na.rm = TRUE)) |>
  arrange(med) |>
  pull(final_ranger_action)

park_model_action <- park_model |>
  mutate(final_ranger_action = factor(final_ranger_action, levels = action_order))

ggplot(park_model_action, aes(x = final_ranger_action, y = duration_hours)) +
  geom_violin(trim = FALSE, fill = "#BFCDE0", color = "#5D7CA6",
              alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.5, fill = "white", outlier.alpha = 0.3) +
  coord_flip() +
  labs(
    title = "Distribution of Response Duration by Final Ranger Action",
    x = "Final Ranger Action (sorted by median duration)",
    y = "Duration of Response (hours)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )

```

The combined violin-and-boxplot visualization provides insight into how long Urban Park Rangers spend responding to different types of animal-related incidents. 

1. Actions Involving Rehabilitation Have the Longest Response Durations

Final actions such as “Released back into Park after Rehabilitation” and “Rehabilitator” show the highest median response durations, as well as the widest spread of times.

These actions often require: on-site assessment, coordination with licensed wildlife rehabilitators, transport of animals, or extended monitoring before the animal can be safely released.

As a result, longer rescue times are expected. The long right-tails in these distributions reflect occasional cases that require many hours of handling or follow-up visits.

2. Administrative or Low-Intervention Actions Are Generally Short

Actions such as: “Advised/Educated others”, “Submitted for DEC Testing”, “Relocated/Condition Corrected”, “Monitored Animal” show much shorter median durations.

These categories often represent cases where: the ranger provides guidance to the public, conducts a quick field assessment, performs minor relocation, or verifies that an animal is behaving normally.

Because they involve limited physical intervention, these responses tend to be resolved quickly, reflected in the tight, compact violins and small IQRs.

3. “Unfounded” and “ACC” Responses Show High Variability

ACC (Animal Care Center Transfer): 
While the median duration is moderate, the long right tail indicates that some cases require significant time—often due to: waiting for ACC personnel to arrive, multi-agency coordination, or handling distressed or dangerous animals.

Unfounded:
Although typically short, several cases show unusually long durations. These may correspond to: extended attempts to locate an animal that was ultimately not found, misreported incidents requiring prolonged investigation, or delays in confirming that no intervention was needed.

4. Outliers Reflect a Small Number of Long-Duration Incidents

Across multiple actions—especially ACC, Unfounded, and Rehabilitator—there are visible outliers extending to 30+ hours. These are likely: long-term monitoring cases, situations requiring multiple returns to the site, or logistical delays in handing off animals to rehabilitation partners.


## Response Duration by Animal Condition

```{r, fig.width=6, fig.height=4, fig.align = "center"}
park_model_cond <- park_model |>
  mutate(
    animal_condition = fct_reorder(
      animal_condition,
      duration_hours,
      .fun = median,
      na.rm = TRUE
    )
  )

ggplot(park_model_cond, aes(x = animal_condition, y = duration_hours)) +
  geom_violin(trim = FALSE, fill = "#BFCDE0", color = "#5D7CA6",
              alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.5, fill = "white", outlier.alpha = 0.3) +
  coord_flip() +
  labs(
    title = "Distribution of Response Duration by Animal Condition",
    x = "Animal Condition",
    y = "Duration of Response (hours)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )
```

The violin-and-boxplot visualization shows how response time varies across different animal conditions. 

1. Unhealthy and Injured Animals Tend to Receive Longer Responses

The Unhealthy and Injured categories show slightly higher median response durations and wider distributions.

This pattern is expected because these cases usually require detailed on-site assessment, safe handling of distressed animals, and possibly coordination with rehabilitation partners.

2. Healthy Animals Show Shorter and More Consistent Response Times

The Healthy category has a shorter median response duration and a narrower distribution.
Healthy animals often require quick verification that no intervention is needed, or simple relocation or observation.

3. Unknown Conditions Have High Variability

The Unknown group displays substantial variation, including several long-duration outliers.
These likely reflect situations where limited information during the call required extended investigation, or the ranger spent additional time determining whether intervention was necessary.

4. DOA Cases Involve Minimal Handling but Occasionally Long Investigations

The DOA (Dead on Arrival) category generally shows short median durations, consistent with the limited intervention required.
However, some notable long-duration outliers appear—possibly tied to: coordinating with external agencies (e.g., ACC), species identification procedures, or waiting for specialized disposal assistance.


## Distribution of Response Duration by Action Group
```{r}
park_model_group = park_model |>
  mutate(
    action_group = if_else(
      final_ranger_action == "Unfounded",
      "Unfounded",
      "Other actions"
    )
  )

duration_mean = park_model_group |>
  group_by(action_group) |>
  summarise(
    mean_duration = mean(duration_hours, na.rm = TRUE),
    .groups = "drop"
  )
duration_mean |>
  knitr::kable(
    digits = 3,
    caption = "Mean Duration by Action Group",
    align = "c"
  )
```


### Density of Raw Response Duration
```{r, fig.width=5, fig.height=3, fig.align = "center"}
ggplot(park_model_group, aes(x = duration_hours,
                        fill  = action_group,
                        color = action_group)) +
  geom_density(alpha = 0.3, adjust = 1, linewidth = 1) +
  
  geom_vline(data = duration_mean,
             aes(xintercept = mean_duration, color = action_group),
             linewidth = 1.2) +
  
  labs(
    title = "Density of response duration by action group",
    x = "Duration of response (hours)",
    y = "Density",
    fill  = "Action group",
    color = "Action group"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )


```

The first density plot shows that the distribution of raw response duration is extremely right-skewed, with most incidents resolved within a short time but a small number taking many hours. This heavy skew makes it difficult to visually compare “Unfounded” cases with all other ranger actions on the original scale because the long tail compresses the main density near zero.

### Density of Log-Transformed Response Duration
```{r, fig.width=5, fig.height=3, fig.align = "center"}
park_model2_log = park_model_group |>
  mutate(duration_log = log1p(duration_hours))  # log(1 + x)

ggplot(park_model2_log,
       aes(x = duration_log, fill = action_group, color = action_group)) +
  geom_density(alpha = 0.4, adjust = 1, linewidth = 0.7) +
  labs(
    title = "Density of log-transformed response duration",
    x = "log(1 + duration of response, hours)",
    y = "Density",
    fill  = "Action group",
    color = "Action group"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

To address this issue, we applied a log-transformation, using log(1 + duration).

This transformation compresses long-duration outliers and spreads out the dense mass near zero, making group differences more interpretable and improving visualization.

Comparison Between Unfounded and Other Actions:

1. Unfounded cases

Display a higher density at lower log-duration values, indicating that these calls are typically resolved very quickly.

The mean duration of unfounded responses is approximately 1.0 hour, confirming their shorter nature.

2. Other actions

Show a broader distribution with higher density at larger log-duration values, meaning that these incidents often require more time for assessment, handling, or coordination.

Their mean duration is significantly higher at 1.52 hours.

## Final Ranger Actions by Animal Condition

```{r}
park_model4 <- park_model |>
  filter(animal_condition %in% c("Healthy", "Unhealthy", "Injured", "DOA"))
```

```{r, fig.align = "center"}
plot_condition_bar <- function(data, cond, ymax) {

  df <- data |>
    filter(animal_condition == cond) |>
    count(final_ranger_action) |>
    arrange(n)

  df$final_ranger_action <- factor(df$final_ranger_action,
                                   levels = df$final_ranger_action)

  ggplot(df, aes(x = final_ranger_action, y = n)) +
    geom_col(fill = "#4F81BD", alpha = 0.85) +
    coord_flip() +
    scale_y_continuous(limits = c(0, ymax),
                       expand = expansion(mult = c(0, 0.05))) +
    labs(
      title = paste("Final Ranger Action for", cond),
      x = "Final Ranger Action",
      y = "Number of Cases") +
    theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.y = element_text(size = 8))
}
max_cases <- park_model4 |>
  count(animal_condition, final_ranger_action) |>
  pull(n) |>
  max()

p1 <- plot_condition_bar(park_model4, "Healthy",   max_cases)
p2 <- plot_condition_bar(park_model4, "Unhealthy", max_cases)
p3 <- plot_condition_bar(park_model4, "Injured",   max_cases)
p4 <- plot_condition_bar(park_model4, "DOA",       max_cases)

(p1 | p2) /
(p3 | p4)
```

The four-panel visualization compares how ranger actions differ depending on the animal’s condition at the time of the call. 

1. Healthy Animals

For Healthy animals, the most common actions are: relocated / condition corrected, ACC.

These results suggest that many healthy animals were either moved from unsafe locations or transferred to the Animal Care Center for further evaluation.
Lower-frequency actions (e.g., rehabilitation) reflect that healthy animals rarely require intensive care.

2. Unhealthy Animals

In the Unhealthy group, ACC overwhelmingly dominates as the most frequent action, followed by Rehabilitator.

Because unhealthy animals typically need medical evaluation or long-term care, leading to high ACC involvement.
Minimal counts in actions like “Submitted for DEC Testing” or “Relocated” indicate that relocation alone is usually insufficient for visibly unhealthy cases.

3. Injured Animals

For Injured animals, the dominant actions are rehabilitator, unfounded, ACC.

Injured animals frequently require hands-on treatment or rehabilitation. 
The relatively high number of “Unfounded” cases likely reflects situations where the animal could not be located or its condition improved by the time the ranger arrived.

4. DOA Animals

Among DOA cases, the most common action is relocated / condition corrected.

This is because dead wildlife generally require removal from public spaces rather than medical intervention.
The presence of some actions like ACC transfer or DEC Testing reflects cases where the animal needed identification or documentation before final handling.




# Additional analysis: If you undertake formal statistical analyses, describe these in detail

## Response Duration

When examining animal condition and borough individually in the raw distributions, neither appears to be a strong or consistent determinant of response duration. These unadjusted patterns suggest that other confounding factors (e.g., call timing, call source) may play a larger role. To evaluate these effects while holding other variables constant, we fit a multiple linear regression model that incorporates incident characteristics including borough, call source, species status, animal condition, age category, animal class, number of animals, and call time.

```{r}
model = lm(response_hrs ~ borough + call_source + species_status + animal_condition + 
             adult + juvenile + infant + animal_class_new + number_of_animals + 
             call_hour_cat, data = ranger_model)

summary(model)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  dplyr::rename(
    Estimate   = Estimate,
    Std_Error  = `Std. Error`,
    t_value    = `t value`,
    p_value    = `Pr(>|t|)`
  ) %>% 
  mutate(
    p_value = format.pval(p_value, digits = 2, eps = 2e-16)
  ) %>% 
  knitr::kable(format  = "markdown", digits = 4, 
               caption = "Regression Coefficient Table", 
               align = c("l", "r", "r", "r", "r")) 
```

We use a multiple linear regression model to estimate the independent contribution of each predictor to response duration, allowing us to isolate the effect of each variable while holding other factors constant.

After adjustment, several predictors show meaningful statistical associations:

**Call time of day** (reference = 8:00-16:00)

- Overnight incidents (0:00–8:00) take about 7.5 hours longer to receive a response, which is one of the strongest effects in the model.

- Evening incidents (16:00–24:00) also experience significantly slower responses, with delays of nearly 4 hours compared to daytime.

- Overall, daytime remains the period with the fastest and most efficient ranger responses. 

- These patterns suggest a progressive decline in responsiveness after normal working hours, likely reflecting reduced staffing and slower processing outside regular working shifts.

**Call source** (reference = Public)

- Incidents reported from Central Dispatch, conservancies, employees, and rangers tend to receive responses that are 1.7–2.2 hours faster than public reports, with the Central Dispatch and employee categories showing statistically significant effects. This suggests that internal reporting channels, which are more direct and reliable, are easier to prioritize, whereas public reports can vary widely in urgency and clarity, potentially slowing response duration.

- Reports categorized as "Other" tend to be slower than public calls, indicating greater variability in information quality or urgency.

- Categories with very small sample sizes (such as WINORR or WBF) display large estimated effects but also uncertainty, so these results should be interpreted cautiously.

**Species status** (reference = Native)

- Incidents involving exotic species require significantly more response duration, with responses roughly 7 hours slower than those involving native species. This delay may reflect additional safety assessments, limited availability of specialized handlers, or the need for species identification before dispatch.

- Other species-status categories show little or no meaningful differences in response duration compared to native species.

**Animal class** (reference = Birds / Raptors)

- Reptile or amphibian cases are handled about 3 hours faster than birds or raptors, which represents a clear and consistent pattern, suggesting these cases require less pre-arrival preparation, whereas birds/raptors may require nets, protective gear, or specialized carriers.

- Fish-related incidents tend to be much slower, but this pattern is unstable due to very small sample size.

- Other animal classes show minor or uncertain differences relative to birds or raptors.

**Borough effects** (reference = Manhattan)

- After adjusting for call characteristics and operational factors, the borough effects differ from the raw patterns observed earlier.

- Compared with Manhattan, both Brooklyn and the Bronx show meaningfully faster response duration (about 1.7–1.8 hours shorter) - although difference between the Bronx and Manhattan is only borderline significant. A possible explanation is that these boroughs may have more concentrated ranger coverage and experience less congestion, allowing for faster travel times.

- In contrast, Staten Island, despite being geographically smaller and less densely populated, shows a possible tendency toward slower responses, but the estimate is not statistically significant. This pattern likely reflects limited ranger staffing, longer travel distances between regions, and the need to cross bridges, which can offset the potential advantages of a smaller service area.

- Queens differs very little from Manhattan, with only a small and negligible difference in response duration

- Overall, borough effects seem to be driven by a combination of coverage density, traffic patterns, and logistical constraints rather than size alone.

In contrast, the following variables do not show meaningful associations with response duration when controlling for other factors:

**Animal Condition** (reference = Healthy)

- Dead-on-arrival cases tend to be handled about 1.5 hours faster, likely because they require fewer preparation, though this difference is not statistically significant.

- Cases involving injured, unhealthy, or unknown animal conditions show small and inconsistent differences compared with healthy animals.

**Age** (Adult, Juvenile, Infant)

- Although the estimated coefficients are relatively large, the effects are not statistically meaningful and do not indicate a real influence on response duration

- These effects likely reflect case-specific variability rather than systematic prioritization.

**Number of Animals**

- The number of animals involved has no meaningful impact on response duration. 

Taken together, these results suggest that operational factors-such as when, where, and through which channel an incident is reported-explain most of the observed differences in response duration. Biological characteristics of the animals, with the exception of exotic species, play a comparatively smaller role. 

```{r}
broom::tidy(model, conf.int = TRUE) %>% 
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Predictor Effects on Response Duration",
       x = "Predictor Variables",
       y = "Estimated Coefficient (hours)") +
  theme_minimal()
```

This coefficient plot highlights the same general patterns seen in the regression table:

- Time of day, call source, exotic species, and several animal classes stand out with clearer effects, whereas most predictors lie near zero or have confidence intervals crossing zero, indicating little or no meaningful association.

Elastic Net was used as a complementary approach to assess variable stability and identify predictors with more consistent signals beyond the Ordinary Least Squares (OLS) estimates.

```{r}
X = model.matrix(~ borough + call_source + species_status + animal_condition + 
                   adult + juvenile + infant + animal_class_new + number_of_animals + 
                   call_hour_cat, data = ranger_model)[, -1]
y = ranger_model$response_hrs

set.seed(123)
fit_cv = cv.glmnet(X, y, alpha = 0.5)

coef(fit_cv, s = "lambda.min") %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "variable") %>% 
  rename(coefficient = 2) %>% 
  filter(coefficient != 0) %>% 
  knitr::kable(format  = "markdown",
    caption = "Elastic Net Coefficients at lambda_min (0.1811)",
    digits = 4,
    align = "l")
```

Compared with the original linear model, the Elastic Net model resulted in a more simplified set of predictors. Several variables that had non-zero coefficients under OLS were reduced to zero, such as `boroughQueens`, `call_sourceWBF`, `call_sourceWINORR`, `species_statusDomestic`, `species_statusUnknown`, `animal_conditionUnknown`, `juvenile` and `infant` age indicators, and `number_of_animals`. This suggests that these variables do not provide stable or meaningful predictive contribution once penalization is applied.

Among the variables that remained, all effects continued to move in the same direction as in the OLS model. The strongest and most consistent patterns also remained largely unchanged: overnight and evening reports still show longer response duration compared with daytime reports. In addition, internal call sources (e.g., Central Dispatch, employees, rangers) continue to be associated with faster responses than public reports. Exotic species consistently require much longer response duration than native species. Certain animal classes, especially reptiles/amphibians (faster) and fish (slower), also retain relatively large effects even after shrinkage.

Several predictors that originally appeared moderately influential in the OLS model, such as the `adult` age indicator, `animal_class_newMarine Mammals`, and `animal_class_newFish`, showed noticeable reductions in magnitude under Elastic Net.  Although the direction of these effects remained the same, their shrinkage suggests that the larger OLS estimates were likely inflated by noise or multicollinearity rather than reflecting a stable underlying effect.

The remaining predictors still show only small or negligible effects in the Elastic Net model, indicating that their contribution to response duration is minimal once penalization removes unstable signals.

Because Elastic Net does not provide p-values or standard errors, these coefficients should be interpreted as indicators of relative importance rather than formal statistical significance.

Overall, response duration appears to be driven primarily by operational characteristics, most notably call timing and call source, rather than biological or borough-specific features. Adjusted modeling reveals patterns not visible in the raw data, highlighting the importance of accounting for confounding factors when interpreting response-time differences.

## ANOVA Models for Response Duration

### Using Raw Duration
```{r}
anova_fit = aov(
  duration_hours ~ final_ranger_action * animal_condition + num_animals,
  data = park_model
)

anova_raw_tbl <- broom::tidy(anova_fit) |>
  select(
    term,
    Df      = df,
    `Sum Sq`  = sumsq,
    `Mean Sq` = meansq,
    `F value` = statistic,
    `Pr(>F)`  = p.value
  )

knitr::kable(
  anova_raw_tbl,
  digits  = 3,
  caption = "ANOVA table for raw response duration",
  align   = "c"
)
```

The original ANOVA model: 
duration_hours ~ final_ranger_action * animal_condition + num_animals

It shows:

Final ranger action is highly significant (p < 2e-16), indicating strong differences in response duration across action types.

Animal condition is not significant (p = 0.46) in the raw model.

Number of animals has no effect (p = 0.80).

The interaction between final action and condition is significant (p = 0.000165), suggesting that the effect of action type depends on the condition of the animal.


### Using Log-Transformed Duration
```{r, fig.align = "center"}
park_model2 = park_model |>
  mutate(log_duration = log(duration_hours))

anova_fit_log = aov(
  log_duration ~ final_ranger_action * animal_condition + num_animals,
  data = park_model2
)

anova_log_tbl <- broom::tidy(anova_fit_log) |>
  select(
    term,
    Df      = df,
    `Sum Sq`  = sumsq,
    `Mean Sq` = meansq,
    `F value` = statistic,
    `Pr(>F)`  = p.value
  )

knitr::kable(
  anova_log_tbl,
  digits  = 3,
  caption = "ANOVA table for log-transformed response duration",
  align   = "c"
)

par(mfrow = c(1, 2))
plot(anova_fit_log, which = 1)  # Residuals vs Fitted
plot(anova_fit_log, which = 2)  # Normal Q-Q
par(mfrow = c(1, 1))


```

Because response duration is highly right-skewed with many long-tail outliers, ANOVA assumptions were violated when using raw duration values. This motivated the use of a log-transformation to stabilize variance and improve normality of residuals.

Model: log_duration ~ final_ranger_action * animal_condition + num_animals

The results:

Final ranger action remains highly significant (p < 2e-16), confirming strong differences in rescue duration across action types even after transformation.

Animal condition becomes significant (p = 0.000416), suggesting that after correcting skewness and long tails, condition differences emerge more clearly.

Number of animals remains non-significant (p = 0.27).

Interaction term remains significant (p = 7.3e-07), indicating that the influence of action types varies across conditions.

The log-transformation compresses extreme high-duration responses and spreads out low values, resulting in more homogeneous variance across fitted values , more normal residual distribution, clearer signal for predictors previously overshadowed by extreme right-skewed values.

This is evident in the diagnostics:

Residuals vs fitted plot shows more stable variance.

Q–Q plot is much closer to the reference line, though long-tail deviations still exist (expected with field-based rescue data).


## Chi-Square Test Between Species Status and Animal Condition
chi-square test between species status and animal condition

```{r}
# conduct a chi-square test on species_status and animal_condition
spec_health_tbl = table(park_ranger_new$species_status, park_ranger_new$animal_condition)

spec_health_result = chisq.test(spec_health_tbl)

# show contingency table of the chi-square test
kable(
  as.data.frame.matrix(spec_health_result$observed),
  caption = "Observed Counts for Species Status and Animal Condition",
  digits = 2
)

# show result tables of the chi-square test
kable(
  as.data.frame.matrix(spec_health_result$expected),
  caption = "Expected Counts for Species Status and Animal Condition",
  digits = 2
)
```

A chi-square test of independence was conducted to evaluate the association between species status and animal condition. The test produced a highly significant result, $$\chi^2(16) = 757.19,\quad p < 2.2 \times 10^{-16}$$, indicating that the distribution of animal conditions differs substantially across species status categories.

The tables of observed and expected counts are shown above, along with the plot of standardized residuals for the two variables. Based on both the contingency table and the residual plot, domestic species are noticeably overrepresented in the Healthy category, while native (non-domestic) species are underrepresented. Domestic species are also underrepresented in the Injured category, whereas native species show the opposite pattern, with more Injured or Unhealthy cases than expected.

# Discussion: What were your findings? Are they what you expect? What insights into the data can you make?

## Species Status and Animal Condition
One possible explanation of the overrepresented healthy cases for domestic species is reporting behavior. Domestic animals may be reported for a wider range of reasons, such as being lost, found unattended, or involved in non-medical encounters, which increases the number of Healthy reports. In contrast, native wildlife is more likely to be reported only when something is wrong, such as being injured, sick, or otherwise in distress, leading to higher counts in those categories.

## Borough and Animal Class
Manhattan has the highest total number of cases, driven largely by small mammals. Manhattan, Queens, and Brooklyn all report particularly high counts for birds/raptors, whereas Staten Island shows the highest numbers for marine mammals, fish, and reptiles/amphibians, and the Bronx reports the highest number of cases involving large mammals.

These patterns may be influenced by differences in human population density and land use. Manhattan, Queens, and Brooklyn have denser human activity, which may increase both encounter rates and incident reporting, especially for small mammals. Human activities in Staten Island often take place near coastal or aquatic environments, making marine animals and other water-dependent species more visible to the public. The Bronx includes the Bronx Zoo and several large parks, which may contribute to a greater presence of large mammals.

Based on these observations, the Parks and Recreation Department could consider allocating more aquatic-animal expertise to Staten Island and strengthening monitoring capacity for its land areas. Additional resources could also be directed toward managing large mammals in the Bronx and reducing human–wildlife disturbances involving small mammals in Manhattan.

